Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-17 paper.tex
Some updates on intervals code.

% - Distribution of breakpoint intervals
%     Study the breakpoint intervals in the Wide ARG. Because we are interested
% in the intervals, we want to be sure that these are robustly reported by the
% HMM. Hence, we focus on those that have two parents in the mirrored and
% unmirrored HMM runs, the same set of mutations, and the left and right parents
% are consistently imputed for Pango lineage status.
%     We focus on the Wide ARG here because it covers (roughly?) the same time
% period as the RIPPLES analysis.

We explore recombination signals across the SARS-CoV-2 genome by inspecting the
breakpoint intervals of the filtered recombination nodes from the wide ARG
(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}. The length
of a breakpoint interval reflects uncertainty in the estimated location of a
breakpoint. On average, the length of the breakpoint intervals is 2,895 bases
(median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of
interval length indicates that the location of a breakpoint often could not be
precisely determined given the observed genetic variation.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1769 breakpoints associated
with 1522 recombination nodes $>1$ descending samples plotted along the genome
as line segments, coloured by interval width. The inset histogram shows the
distribution of these interval widths.
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
more closely, we find an enrichment in the number of intersecting intervals in
the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
4B). Markedly, there is an abrupt drop in the number of intersecting intervals
in the N gene following the peak in ORF8 towards the 3’ end.

@@

The remaining remaining 630 Pango recombinant samples, representing 31
Nextclade-designated Pango lineages, trace back to 50 different most recent
recombination nodes.
We examine the inferred inheritance structure in the following subsections.

@@

\subsubsection{Single origin}
In the absence of genealogical information, a reasonable initial assumption is
that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. We would expect our ARG to reveal evolutionary histories of this
nature, where all the samples assigned to a given recombinant lineage
trace back to one recombination node, representing a single originating
recombination event.

@@

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node use
breakpoint positions in the ARG, corresponding to the maximum inferred breakpoint position
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show other descendants of recombinant parents. It is clear
from the subgraph that all the XA
samples trace to a single originating recombination node,
whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.

@@

A more complex single-origin case is XAG, illustrated in
Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
all trace back to the same most recent common recombination node
(combining BA.1 on the left and BA.2.9 on the right), but
we infer this recombination event to also be the originating event for
all the recombinant samples designated XAA,
and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.

@@

In total out of the 31 Pango recombinant lineages in the Long ARG we identify 16 as
having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations.

\subsubsection{Multiple origins}
The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine several Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e. trace
back to independent recombination nodes in the ARG).

@@

A similar pattern is seen for several of the origins inferred for the Pango
XE lineage (originally defined on the basis of sequences from the UK and Ireland). The
Long ARG infers five independent origins for nonsingleton Nextclade-designated XE
samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
(primarily UK and Irish, but with a few northern European and north American strains).
This certainly represents the XE recombinant lineage as originally defined. The other recombination nodes
only give rise for one, two, or three XE samples. Three of these recombination
nodes are likely to represent Nextclade mislabelling: they either have different
inferred breakpoint positions or (in the case of the earliest recombination node),
descendants that are primarily labelled as other Pango X-lineages. Additional support
for the suggestion that they result from incorrect labelling comes from the fact that
none of their XE-labelled descendant samples are from the UK or Ireland.
It is less clear if the most recent XE cluster is a mislabelling or a case of \texttt{sc2ts}
inferring multiple recombinations when there should be just one. This cluster of 2 samples
(one of which is English) has exactly the same breakpoint estimates as the
main cluster. If the only samples available are rather genetically distant from an originating
recombinant, we might expect sc2ts to infer multiple similar recombination events
(rather than a single event with extra reversions)

@@

\subsubsection{Nested recombinant origins}
As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, four of the
Pango X lineages (XB, XE, XAB, and XAJ) have Nextclade-designated samples which
are related in a more complex manner, where multiple non-independent (``nested'')
recombination events occur. An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Similarly, XE and XAJ
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Also note that the number of mutations on the lineages immediately above and
below the recombination node (totalling 18+2) is rather large, suggesting
that the sampled recombinant which induced the recombination node in the Long ARG
is only distantly related to the true originating recombinant. This could account
for complex and potentially artefactual relationships around these nodes, and
is likely to be due to undersampling of the XB outbreak. Investigating examples
of nested recombinant origins, and identifying which (if any) of the nested
recombination events may be artefactual, is an important area of future research.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node. The samples descending from these nested
recombinations include 9 that are assigned by Nextclade to
various non-recombinant pre-Alpha lineages (blue).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/long_arg_recombination_intervals.pdf}
\caption{\label{fig:long_arg_breakpoint_distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Long ARG.}
\end{figure}

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: other descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-single-origin}
Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text). In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
are taken from Pango designation alias key (TODO: explain the asterisks). Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
with 1,522 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width). The inset histogram shows the
distribution of these interval widths.
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
The top-ten sites by mutation count are annotated.
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
intervals and patterns of recurrent mutation along the genome in the Wide ARG
(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
of the results. To reduce the effect of artefactual recombinants, we
consider only the breakpoints associated with the 1,522 recombination nodes
that are ancestral to more than one sample. The mean length of these intervals
is 1,685 bases (median 962); the length distribution is summarised in the
inset histogram in Figure~\ref{fig:breakpoint-distribution}.

Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
Other than the boundaries happen to lie on gene breaks, not an awful lot I
think?]

% This isn't really true any more, is it?
% There is a preponderance of breakpoint intervals spanning the 3’ end of the
% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
% more closely, we find an enrichment in the number of intersecting intervals in
% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
% in the N gene following the peak in ORF8 towards the 3’ end.

@@

The remaining remaining 630 Pango recombinant samples trace back to 50 different
most recent recombination nodes, representing 31
Nextclade-designated Pango lineages. These are summarised
in Supplementary Table~\ref{tab:pango-recombinants}, and fall
into three classes: single origin, multiple origin,
and multiple nested origins.

@@

\subsubsection{Single origin}
In the absence of genealogical information, a reasonable initial assumption is
that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. We would expect our ARG to reveal evolutionary histories of this
nature, where all the samples assigned to a given recombinant lineage
trace back to one recombination node, representing a single originating
recombination event. In the Long ARG, 16 of the 31 Pango recombinants lineages identified by Nextclade
 fall into this category (Supplementary Table~\ref{tab:pango-recombinants})

@@

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

One of the simplest examples is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
Here, a subgraph of the long ARG has been plotted, where paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show additional descendants of recombinant parents.

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a single originating recombination node,
whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.

@@

A more complex single-origin case is XAG, illustrated in
Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
all trace back to the same most recent recombination node
(combining BA.1 on the left and BA.2.9 on the right), but
we infer this recombination event to also be the originating event for
all the recombinant samples designated XAA,
and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.

@@

Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type with no other Pango designations among their descendants. 
The remaining 10 are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations. Supplementary Table~\ref{tab:pango-recombinants} also shows the official Pango designated parent lineages for
each recombinant lineage. The parent lineages in the Long ARG agree to a great extent with these, although \texttt{sc2ts} provides more a precise
parent designation.

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to >1 most recent
recombination node in the ARG. Of these, 11 are cases where none of the
recombinants are nested (i.e. there is no overlap between the descendant samples
of each recombination node). In 9 of these 11, there is a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (in other two, there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

@@

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). The
Long ARG infers four independent origins for nonsingleton Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
(primarily UK and Irish, but with a few northern European and north American strains).
This certainly represents the XE recombinant lineage as originally defined. The other recombination nodes
only give rise for one, two, or three XE samples. Three of these recombination
nodes are likely to represent Nextclade mislabelling: they either have different
inferred breakpoint positions or (in the case of the earliest recombination node),
descendants that are primarily labelled as other Pango X-lineages. Additional support
for the suggestion that they result from incorrect labelling comes from the fact that
none of their XE-labelled descendant samples are from the UK or Ireland.
It is less clear if the most recent XE cluster is a mislabelling or a case of \texttt{sc2ts}
inferring multiple recombinations when there should be just one. This cluster of 2 samples
(one of which is English) has exactly the same breakpoint estimates as the
main cluster. If the only samples available are rather genetically distant from an originating
recombinant, we might expect sc2ts to infer multiple similar recombination events
(rather than a single event with extra reversions)

@@

\subsubsection{Nested recombinant origins}

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A subgraph of the Long ARG showing
nested recombination events involving Pango lineage XB. All XB samples
trace to a single recombination node (top centre), but three further recombinations
also descend from this node. The samples descending from these nested
recombinations include 9 that are assigned by Nextclade to
various non-recombinant pre-Alpha lineages (blue).}
\end{figure}

As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-single-origin}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

Note that in the Long ARG, the nested recombination events account for only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Also note that the number of mutations on the lineages immediately above and
below the recombination node (totalling 18+2) is rather large, suggesting
that the sampled recombinant which induced the recombination node in the Long ARG
is only distantly related to the true originating recombinant. This could account
for complex and potentially artefactual relationships around these nodes, and
is likely to be due to undersampling of the XB outbreak. Investigating examples
of nested recombinant origins, and identifying which (if any) of the nested
recombination events may be artefactual, is an important area of future research.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/long_arg_recombination_intervals.pdf}
\caption{\label{fig:long_arg_breakpoint_distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Long ARG.
Top panel shows the intervals for 851 breakpoints associated
with 763 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width).
Other details as described in Figure~\ref{fig:breakpoint-distribution}.}
\end{figure}

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-recombinants}
Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text). In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
are taken from Pango designation alias key (TODO: explain the asterisks). Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}
